$p$-adic precision

### 2014-02-15 padicprec.tex
Comments on the non-surjectivity of the differential.

From the beggining, we have assumed that $f'(x)$ is surjective. Let us 
discuss shortly what happens when this assumption is relaxed. As it is 
explained after the proof of Lemma \ref{lem:main}, the first thing we 
can do is to project the result onto different quotients, \emph{i.e.} to 
work with the composites $\pr_K \circ f$ for a sufficiently large family 
of closed sub vector spaces $K \subset F$ such that $K + f'(x)(E) = F$. 
(We recall that here $\pr_K$ denotes the canonical projection to $F$ 
onto the quotient $F/K$.) If $F$ has a natural system of coordinates 
(which is the case in almost all settings), we can generally take the 
$\pr_K$'s as the projections on each coordinate. Doing this, we end up 
with a precision on each individual coordinate. Furthermore, we have the 
guarantee that all these precisions are sharp (even if the lattice build
from them is not).
Let us illustrate the above discussion by an example: suppose that we 
want to compute the function $f : (K^n)^n \to M_n(K)$ that takes a 
family of $n$ vectors to its Graam matrix. The differential of $f$ is
clearly never surjective because $f$ takes its values in the subspace
consisting of symmetric matrices. Nevertheless for all couple $(i,j)
\in \{1, \ldots, n\}^2$, one can consider the composite $f_{ij} = 
\pr_{ij} \circ f$ where $\pr_{ij} : M_n(K) \to K$ that takes a matrix 
$M$ to its $(i,j)$-th entry. The maps $f_{ij}$'s are differentiable and 
their differentials generally are surjective (\emph{i.e.} they do not 
vanish). Let $M$ be a matrix known at some finite precision such that 
$f'_{ij} (M) \neq 0$ for all $(i,j)$. We can then apply the forward 
and/or the backward computation and get $f_{ij}(M)$ together with its 
precision. Putting this together, we get the whole matrix $f(M)$ 
together with a sharp precision datum on each entry.

The study of the this example actually suggests another solution to 
tackle the issue of non-surjectivity. Indeed, remark that our $f$ above 
had not a surjective differential simply because its codomain was too 
large: if we had replaced $f : (K^n)^n \to M_n(K)$ by $g : (K^n)^n \to 
S_n(K)$ (where $S_n(K)$ denotes the $K$-vector space of symmetric matrix 
over $K$ of size $n$) defined in the same way, our problem would 
disappear. Of course the image of a general $f$ is rarely a sub vector 
space of $F$ but it is often a sub-$K$-manifold of $F$. 
% Using the results of Appendix

@@

We conclude this Section by remarking that the two strategies designed 
above (\emph{i.e.} forward and backward computations) share some 
similarities with usual floating point arithmetics over the reals. 
Indeed, roughly speaking, in each setting, we begin by choosing a large 
precision, we do all our computations up to this precision understood 
that when we are not sure about some digit, we choose it ``at random'' 
or using good heuristics. The main difference is that, in the 
ultrametric setting, we are able (under some mild hypothesis) to 
quantify the precision we need at each individual step in order to be 
sure that the final result is correct up to the required precision.

>>>>

From the beggining, we have assumed that $f'(x)$ is surjective. Let us 
discuss shortly what happens when this assumption is relaxed. As it is 
explained after the proof of Lemma \ref{lem:main}, the first thing we 
can do is to project the result onto different quotients, \emph{i.e.} to 
work with the composites $\pr_K \circ f$ for a sufficiently large family 
of closed sub vector spaces $K \subset F$ such that $K + f'(x)(E) = F$. 
(We recall that here $\pr_K$ denotes the canonical projection to $F$ 
onto the quotient $F/K$.) If $F$ has a natural system of coordinates 
(which is the case in almost all settings), we can generally take the 
$\pr_K$'s as the projections on each coordinate. Doing this, we end up 
with a precision on each individual coordinate. Furthermore, we have the 
guarantee that all these precisions are sharp (even if the lattice build
from them is not).

Let us illustrate the above discussion by an example: suppose that we 
want to compute the function $f : (K^n)^n \to M_n(K)$ that takes a 
family of $n$ vectors to its Graam matrix. The differential of $f$ is
clearly never surjective because $f$ takes its values in the subspace
consisting of symmetric matrices. Nevertheless for all couple $(i,j)
\in \{1, \ldots, n\}^2$, one can consider the composite $f_{ij} = 
\pr_{ij} \circ f$ where $\pr_{ij} : M_n(K) \to K$ that takes a matrix 
$M$ to its $(i,j)$-th entry. The maps $f_{ij}$'s are differentiable and 
their differentials generally are surjective (\emph{i.e.} they do not 
vanish). Let $M$ be a matrix known at some finite precision such that 
$f'_{ij} (M) \neq 0$ for all $(i,j)$. We can then apply the forward 
and/or the backward computation and get $f_{ij}(M)$ together with its 
precision. Putting this together, we get the whole matrix $f(M)$ 
together with a sharp precision datum on each entry.

The study of the this example actually suggests another solution to 
tackle the issue of non-surjectivity. Indeed, remark that our $f$ above 
had not a surjective differential simply because its codomain was too 
large: if we had replaced $f : (K^n)^n \to M_n(K)$ by $g : (K^n)^n \to 
S_n(K)$ (where $S_n(K)$ denotes the $K$-vector space of symmetric matrix 
over $K$ of size $n$) defined in the same way, our problem would have 
disappeared. Of course the image of a general $f$ is rarely a sub vector 
space of $F$ but it is often a sub-$K$-manifold of $F$. We can then use 
the results of Appendix \ref{sec:manifold} to study $f$ viewed as a 
function whose codomain is $f(U)$, understood that the differential of 
it has now good probability to be surjective.

@@

We conclude this Section by briefly remarking that the two strategies designed 
above (\emph{i.e.} forward and backward computations) share some 
similarities with usual floating point arithmetics over the reals. 
Indeed, roughly speaking, in each setting, we begin by choosing a large 
precision, we do all our computations up to this precision understood 
that when we are not sure about some digit, we choose it ``at random'' 
or using good heuristics. The main difference is that, in the 
ultrametric setting, we are able (under some mild hypothesis) to 
quantify the precision we need at each individual step in order to be 
sure that the final result is correct up to the required precision.
